Measuring causal influence with\\ back-to-back regression: the linear case

### 2019-05-23 paper/introduction.tex
a

Both forward and backward modeling have competing benefits and drawbacks. Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations. By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \cite{weichwald2015causal, hebart2018deconstructing, king2018encoding}. There also exist cross-decomposition techniques such as Partial Least Squares (PLS) and Canonical Correlation Analysis (CCA), but they are notoriously difficult to interpretat \citep{lebart1995statistique}.

We introduce the 'back-to-back regression', which combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}) After detailing the method and proving its convergence (Section~\ref{sec:theorem}), we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques, both in terms of estimating causal influence and out-of-environment predictions (Section~\ref{sec:experiment_synthetic}). Finally, we apply back-to-back regression to a large neuroimaging dataset and reveal that distinct but collinear word features leads to distinguishable brain responses (Section~\ref{sec:experiment_real}).

>>>>

Both forward and backward modeling have competing benefits and drawbacks. Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations. By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \cite{weichwald2015causal, hebart2018deconstructing, king2018encoding}. There also exist cross-decomposition techniques such as Partial Least Squares (PLS) and Canonical Correlation Analysis (CCA), but they are notoriously difficult to interpretet \citep{lebart1995statistique}.

We introduce the 'back-to-back regression', which combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}) After detailing the method and proving its convergence (Section~\ref{sec:theorem}), we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques, both in terms of estimating i) causal influence and ii) out-of-environment predictions (Section~\ref{sec:experiment_synthetic}). Finally, we apply back-to-back regression to a large neuroimaging dataset and reveal that distinct but collinear word features leads to distinguishable brain responses (Section~\ref{sec:experiment_real}).

### 2019-05-23 paper/paper.tex
Scholkopf
a

Back-to-Back Regression (B2B) consists in two steps.
%
First, we estimate the linear regression coefficients $\hat G$ from $Y$ to $X$, and construct the predictions $\hat X = \hat Y \hat G$.
%
This backward regression recovers an approximation of the causes of $Y$ in $X$.
%
Second, we estimate the linear regression coefficients $\hat H$ from $X$ to $\hat X$.
%
The diagonal of the regression coefficients $\hat H$, denoted by $\hat{E} = \text{Diag}(E)$, is the desired estimate of the causal influence matrix $E$.

@@

This section provides a consistency proof for B2BR in the case of low noise.

\begin{theorem}[Low-noise consistency]
    Consider the B2BR model from Equation~\ref{eq:model}.
    %
    Assume that $F$ and $X$ are full-rank on the subspace spanned by active entries in $E$. 
    %
    Then, as the variance of $N$ tends to zero and the sample size tends to infinity, B2BR recovers $E$.
\end{theorem}
\begin{proof}
For centered noise $N$, the first regression satisfies:
\begin{align*}
    \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2] &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2]\\
                                                        &{}= \arg \min_G \mathbb{E}[\left \| X - XEFG\right\| ^2]  + \mathbb{E}[\left \| NFG\right \| ^2].
\end{align*}
%
The solution to this first regression is $\hat{G} = [(XEF)^\top (XEF) + (NF)^\top (NF)]^{-1} (XEF)^\top X$.

@@

On the one hand, we have leveraged B2B to obtain the estimate $\hat{E}$, which is a diagonal matrix with real entries.
%
On the other hand, the true causal influence matrix $E$ is a binary matrix, which hard-selects the causal factors of $Y$ from $X$.
%
In this section, we provide a recipe to binarize $\hat{E}$ to estimate the collection of causal factors.

@@

\section{Related work}
% scholkopf2016modeling

@@

B2BR is also related, but different, from usual causal discovery algorithms \citep{peters2017elements} that hypothesize on the direction of causality, or from measures of causal influence requiring time information \citep{granger1969investigating, janzing2013quantifying}.

@@

>>>>

Back-to-Back Regression (B2B) consists in two steps.
%
First, we estimate the linear regression coefficients $\hat G$ from $Y$ to $X$, and construct the predictions $\hat X = Y \hat G$.
%
This backward regression recovers the correlations between $Y$ and each factor of $X$.
%
Second, we estimate the linear regression coefficients $\hat H$ from $X$ to $\hat X$.
%
The diagonal of the regression coefficients $\hat H$, denoted by $\hat{E} = \text{Diag}(E)$, is the desired estimate of the causal influence matrix $E$.

@@

This section provides a consistency proof for B2B in the case of low noise.

\begin{theorem}[Low-noise consistency]
    Consider the B2B model from Equation~\ref{eq:model}.
    %
    Assume that $F$ and $X$ are full-rank on the subspace spanned by active entries in $E$.
    %
    Then, as the variance of $N$ tends to zero and the sample size tends to infinity, B2B recovers $E$.
\end{theorem}
\begin{proof}
For centered noise $N$, the first regression satisfies:
\begin{align*}
    \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2] &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2]\\
                                                        &{}= \arg \min_G \mathbb{E}[\left \| X - XEFG\right\| ^2]  + \mathbb{E}[\left \| NFG\right \| ^2].
\end{align*}
%
The solution to this first regression is $\hat{G} = [(XEF)^\top (XEF) + (NF)^\top (NF)]^{-1} (XEF)^\top X$.

@@

On the one hand, we have leveraged B2B to obtain the estimate $\hat{E}$, which is a diagonal matrix with real entries.
%
On the other hand, the true causal influence matrix $E$ is a binary matrix, which hard-selects the causal factors of $Y$ from $X$.
%
In this section, we provide a recipe to binarize $\hat{E}$ and estimate the collection of causal factors.

@@

\section{Related work}

@@

B2B is also related, but different, from usual causal discovery algorithms \citep{peters2017elements} that hypothesize on the direction of causality, or from measures of causal influence requiring time information \citep{granger1969investigating, janzing2013quantifying}.
%
B2B seems to have a similar flavour to half-sibling regression \citep{scholkopf2016modeling}, in the sense that we observe a large number of variables corrupted by the same source of noise; we leave the study of the precise link between these two models for future work.

@@

% \input{supplementary.tex}

### 2019-05-23 paper/synthetic.tex
a

We evaluate the performance of B2BR throughout a series of experiments on controlled synthetic data.
%
The purpose of these experiments is to evaluate the ability of B2BR in terms of prediction (inside and outside the training distribution), as well as a method to recover causal factors.

The data generating process for each experiment constructs $n$ training examples according to the model $Y = (\text{snr} \cdot XE + N)F$.
%
Here, 
%\begin{itemize}
%    \item $F \in \mathbb{R}^{d_x \times d_y}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
%    \item $X \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_X)$,
%    \item $N \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_N)$,
%    \item $E \in \mathbb{R}^{d_x \times d_x}$ is a binary diagonal matrix containing $n_c$ ones,
%    \item $\Sigma_X = AA^\top$, where $A \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
%    \item $\Sigma_N = BB^\top$, where $B \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
%\end{itemize}
    $F \in \mathbb{R}^{d_x \times d_y}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
    $X \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_X)$,
    $N \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_N)$,
    $E \in \mathbb{R}^{d_x \times d_x}$ is a binary diagonal matrix containing $n_c$ ones,
    $\Sigma_X = AA^\top$ where $A \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
    $\Sigma_N = BB^\top$ where $B \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
and the factor $\text{snr} \in (0, \infty)$.

@@

We compare the performance of B2BR against five competing methods:
%
Ridge regression \citep{hoerl1959optimum}, multitask Lasso \citep{argyriou2008convex}, Partial Least Squares or PLS \citep{wold_pls, tenenhaus_pls}, Canonical Correlation Analysis or CCA \citep{cca_hotelling}, and Reduced Rank Regression or RRR \citep{Izenman_rrr}.
%
Each of these methods estimates a matrix of coefficients $\hat{W} \in \mathbb{R}^{d_x \times d_y}$, from which we estimate the diagonal mask $\hat{E}$ using the feature importances $\hat{E}_{i,i} = \| W_{i, :} \|$.
%
For B2BR, we obtain $\hat{E}$ as described in Algorithm~\ref{algorithm:b2br}.
%
We employ $5$-fold cross-validation to select the optimal hyper-parameters of each method.

@@

Table~\ref{table:synthetic} summarizes the results for all experimental conditions, metrics, and methods.
%
B2BR is the method obtaining best test error out-distribution, lowest false positive and false negative discovery about $E$, and overall the best AUC when it comes to detecting causal influences.

>>>>

We evaluate the performance of B2B throughout a series of experiments on controlled synthetic data.
%
The purpose of these experiments is to evaluate the ability of B2B in terms of prediction (inside and outside the training distribution), as well as a method to recover causal factors.

The data generating process for each experiment constructs $n$ training examples according to the model $Y = (\text{snr} \cdot XE + N)F$, where $snr$ is a scalar that modulates the signal-to-noise ratio.
%
Here,
%\begin{itemize}
%    \item $F \in \mathbb{R}^{d_x \times d_y}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
%    \item $X \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_X)$,
%    \item $N \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_N)$,
%    \item $E \in \mathbb{R}^{d_x \times d_x}$ is a binary diagonal matrix containing $n_c$ ones,
%    \item $\Sigma_X = AA^\top$, where $A \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
%    \item $\Sigma_N = BB^\top$, where $B \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
%\end{itemize}
    $F \in \mathbb{R}^{d_x \times d_y}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
    $X \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_X)$,
    $N \in \mathbb{R}^{n \times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_N)$,
    $E \in \mathbb{R}^{d_x \times d_x}$ is a binary diagonal matrix containing $n_c$ ones,
    $\Sigma_X = AA^\top$ where $A \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
    $\Sigma_N = BB^\top$ where $B \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from $\mathcal{N}(0, d_x^{-1})$,
and the factor $\text{snr} \in (0, \infty)$.

@@

We compare the performance of B2B against five competing methods:
%
Ridge regression \citep{hoerl1959optimum}, multitask Lasso \citep{argyriou2008convex}, Partial Least Squares or PLS \citep{wold_pls, tenenhaus_pls}, Canonical Correlation Analysis or CCA \citep{cca_hotelling}, and Reduced Rank Regression or RRR \citep{Izenman_rrr}.
%
Each of these methods estimates a matrix of coefficients $\hat{W} \in \mathbb{R}^{d_x \times d_y}$, from which we estimate the diagonal mask $\hat{E}$ using the feature importances $\hat{E}_{i,i} = \| W_{i, :} \|$.
%
For B2B, we obtain $\hat{E}$ as described in Algorithm~\ref{algorithm:b2br}.
%
We employ $5$-fold cross-validation to select the optimal hyper-parameters of each method.

@@

Table~\ref{table:synthetic} summarizes the results for all experimental conditions, metrics, and methods.
%
B2B is the method obtaining best test error out-distribution, lowest false positive and false negative discovery about $E$, and overall the best AUC when it comes to detecting causal influences.
